---
title: 'Statistical Inference Course Project : Part 2'
author: "Alaettin Serhan Mete"
date: "7/21/2017"
output:
  pdf_document: default
  html_document: default
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8,
                      echo=TRUE, warning=FALSE, message=FALSE)
```

# Basic Inferential Data Analysis

In this part we're going to analyze the ToothGrowth data in the R datasets package.
We're going to use confidence intervals and hypothesis tests to compare tooth growth by supp and dose.
Let's first load the dataset and print some summary information, along with a 
few explatory histograms:

```{r}
# Load the tooth growth data
dataset <- ToothGrowth

# Print some basic summary of the data
print(str(dataset))

# Print some basic summary of the data
print(summary(dataset))
```

```{r}
# Let's quickly look at 1-D histograms of len an dose
hist(dataset$len, 
     main = "Histogram of Tooth Length",
     xlab = "Tooth length",
     ylab = "Frequency",
     ylim = c(0,20), 
     col  = "cyan",
     lwd  = 3)

# Let's quickly look at 1-D histograms of len an dose
hist(dataset$dose, 
     main = "Histogram of Dose",
     xlab = "Dose (milligrams/day)",
     ylab = "Frequency",
     ylim = c(0,25), 
     col  = "magenta",
     lwd  = 3)
```

As one can see there are:

*  Two different supplement types : OJ (orange juice) and VC (vitamin C)
*  Three different dose types : 0.5, 1.0 and 2.0 (milligrams/day)
*  The tooth length varies from 0 to 35 (I'm guessing in units of mm?)

Now, let's make a box plot to see how tooth lenght varies with the supplement type
and dose:

```{r}
# Now plot length~supp
boxplot(len~supp+dose, data = dataset,
        main = "Tooth Growth for Different Supplement Types and Doses",
        xlab = "Supplement Type (OJ, VC) . Dose (0.5, 1, 2)", ylab = "Length",
        col  = c("blue","green"))
axis(side = 2, lwd = 3)
```

In this plot, the color blue corresponds to supplement type OJ and green corresponds to VC.
The distribution is further split into dose types, namely 0.5, 1, and 2 milligram/day. 
Each box shows the 25 and 75 percentiles with the horizantal black line in the middle showing the median.

There are a few basic information one can gather from this plot. 
It seems as if as the dose increases, the tooth length also increases.
Except for the dose of 2 milligrams per day, the median tooth length for the supplementary type OJ
is larger than the supplementary type VC. For the dose of 2 milligrams per day, they are about the same.

Now let's run some tests to quantify these features:

Let's first check if there is a disagreement between supplement types (for all doses combined):

```{r}
# Let's perform t-test
t.test(len ~ supp, paired = FALSE, var.equal = FALSE, data = dataset)
```

As can be seen the 95% CL spans the interval [-0.17, 7.57] including zero, meaning
one cannot reject the null hypothesis (there is no dependence on the supplement type).
Therefore, one can say, with 95% CL, both supplement types result in comparable tooth growth. 

Now let's run some tests to see if there is a correlation between the tooth growth and the supplementary doses
 (for both types combined):

```{r}
# First split the data into dose types:
low.mid.dose  <- subset(dataset, dose %in% c(0.5,1.0))
mid.high.dose <- subset(dataset, dose %in% c(1.0,2.0))
low.high.dose <- subset(dataset, dose %in% c(0.5,2.0))
```

```{r}
# Let's perform t-test for the 0.5 and 1.0
t.test(len ~ dose, paired = FALSE, var.equal = FALSE, data = low.mid.dose)
```

```{r}
# Let's perform t-test for the 1.0 and 2.0
t.test(len ~ dose, paired = FALSE, var.equal = FALSE, data = mid.high.dose)
```

```{r}
# Let's perform t-test for the 0.5 and 2.0
t.test(len ~ dose, paired = FALSE, var.equal = FALSE, data = low.high.dose)
```

In all three tests, we see that none of the 95% CL intervals include zero and they are always 
on the negative side (remember we have the lower dose - higher dose in the test). Therefore,
we can claim, w/ 95% CL, that the tooth growth increases as the supplementary dose increases 
(i.e. they are positively correlated).

Our assumptions on these tests were:

 * Test subjects are randomly assigned type/dose therefore we performed unpaired tests,
 * For each invidual group we didn't assume a common variance (i.e var.equal = FALSE), 
 this could've been assumed otherwise too.

To wrap up, when all doses are combined, we see that there is no strong evidence that suggests that 
the tooth grow depends on the supplement type. Moreover, when supplement types are combined, we see that
there is an evidence that suggests a positive correlation between the amount of dose and the tooth growth
(i.e. the larger the dose the larger the tooth grow).