---
title: 'Statistical Inference Course Project : Part 1'
author: "Alaettin Serhan Mete"
date: "7/21/2017"
output:
  pdf_document: default
  html_document: default
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8,
                      echo=TRUE, warning=FALSE, message=FALSE)
```

# Simulation Exercise

In this part we're going to explore the Central Limit Theorem (CLT) and its implications.
To do this first we're going to perform 1000 simulations, in each iteration we're
going to generate 40 random numbers using an exponential distribution with the rate
parameter of 0.2. 

The true mean of the exponential distribution is 1/rate, so is its 
standard deviation. So, in this case the true mean is 1/0.2 = 5. If we sample the means
1000 times, the mean of the means will also be 5, and its standard error will be
the original standard deviation divided by the square root of the number of random draws,
i.e. in this case standard error of means will be 5/sqrt(40) = 0.79. This distribution is
supposed to look like a normal distribution, at least as claimed by the CLT, even though
the original distribution (exponential) is not. Let's see if this is indeed the case:

```{r}
# Set the seed for reproducibility
set.seed(1234)

# First set the variables as pointed out in the assignment
rate  <- 0.2 
n.exp <- 40
n.sim <- 1e3

# Now compute the true mean/sd and standard error of means (sig/sqrt(n))
true.mean   <- 1/rate
true.sd     <- 1/rate
se.of.means <- true.sd/sqrt(n.exp)

# Now throw 1k toys, generate 40 exp w/ rate 0.2 and store the means and sds
means <- numeric()
sds   <- numeric()
for(i in seq(1,n.sim,by=1)) {
    values <- rexp(n = n.exp, rate = rate)
    means  <- c(means,mean(values))
    sds    <- c(sds,sd(values))
}

# Now let's look at the distribution of the means
hist(means, breaks  = seq(1.75,8.25,by=0.5), 
     main = "Histogram of Means",
     xlab = "Means",
     ylab = "Density",
     ylim = c(0, 0.6),
     prob = TRUE,
     lwd  = 3,
     col  = "orange")

# Draw a line for the true mean and annotate it
abline(v = 1/rate, lwd = 3, col = "blue")
text(x = 1/rate + 0.1, y = 0.25, "True Mean", srt = 90., col = "blue")

# Now let's fit this sample w/ a normal distribution and see how it looks
if(!require(MASS)) {
    install.packages("MASS")
    require(MASS)
}
fittedmeans <- fitdistr(means, "normal")
parameters  <- fittedmeans$estimate

# Draw the fitted gaussian and annotate the fitter values
curve(dnorm(x,parameters[1],parameters[2]), 
      add = TRUE,
      col = "green",
      lwd = 3)
text(x = 7, y = 0.55, 
     paste("Fitted sample mean = ", round(parameters[1],2) , " sd = ", round(parameters[2],2)),
     col = "green", cex = 1.25)

# Now plot the true normal and annotate
curve(dnorm(x,true.mean,se.of.means), 
      add = TRUE,
      col = "blue", 
      lwd = 3)
text(x = 7, y = 0.50, 
     paste("True sample mean = ", round(true.mean,2) , " sd = ", round(se.of.means,2)),
     col = "blue", cex = 1.25)

```

In the above plot, the orange histogram shows the means of the 40 randomly drawn exponentials
in 1000 simulations (i.e. there are 1000 entries in the histogram). The distribution is normalized
to unity area to show the density instead of frequency. The green curve is the fitted normal distribution
with mean of 4.97 and sd of 0.76, while the blue curve shows a normal distribution with the
true mean of 5 and standard error of means 0.79 (5/sqrt(40)). As one can see, they agree very nicely,
showing the CLT actually works :) 
